Data simulation features

### 2017-11-13 inst/gitbook/04_making-up-data.Rmd
Updated manuals

Once a matrix is generated, it is possible to apply inapplicable characters to it for increasing realism!
Inapplicable characters are commonly designated as `NA` or simply `-`.
They differ from missing characters `?` in their nature by being inapplicable rather than unknown.
For example, considering a binary character defined as "color of the tail" with the following states "blue" and "red"; on a taxon with no tail, the character should be coded as inapplicable ("`-`") since the state of the character "color of tail" is *known*: it's neither "blue" or "red", it's just not there!
It contrasts with coding it as missing ("`?`" - also called as ambiguous) where the state is *unknown*, for example, the taxon of interest is a fossil where the tail has no colour preserved or is not present at all due to bad conservation!

@@

## The first few lines of the resulting matrix
my_matrix_NA[1:10, 1:10]
```
<!--TG: Add inapplicable paper here -->

@@

There are many parameters that can create a "realistic" matrix (i.e. not too different from the input tree with a consistency and retention index close to what is seen in the literature) but because of the randomness of the matrix generation not all end up creating "good" matrices.
The following parameters however, seem to generate fairly "realist" matrices with a starting coalescent tree, equal rates model with 0.85 binary characters and 0.15 three state characters, a gamma distribution with a shape parameter ($\alpha$) of 5 and no scaling ($\beta$ = 1) with a rate of 100.

@@

## Simulating multidimensional spaces
Another way to simulate data is to directly simulate an ordinated space with the `space.maker` function.
This function allows users to simulate multidimensional spaces with a certain number of properties.
It takes as arguments the number of elements (data points) and dimensions to create the space but also permits more fine tuning in the data simulation: it is possible to give a specific distribution to each dimension, provide a correlation matrix to link the dimensions or even provide an *a priori* distribution of the variance for each distributions!
<!-- ooh fancy! -->

[@diaz2016global]

@@

## A cube space
scatterplot3d(space.maker(2500, 3, runif), pch = 20,
              xlab = "", ylab = "", zlab = "",
              main = "Uniform cube space")

## A plane space
scatterplot3d(space.maker(2500, 3, c(runif, runif, runif),
              arguments = list(list(min = 0, max = 0), NULL, NULL)), pch = 20,
              xlab = "", ylab = "", zlab = "",
              main = "Uniform plane space")

## An ellipsoid space (=a spheric space with correlation)
cor_matrix <- matrix(cbind(1, 0.8, 0.2, 0.8, 1, 0.7, 0.2, 0.7, 1), nrow = 3)
scatterplot3d(space.maker(2500, 3, rnorm, cor.matrix = cor_matrix), pch = 20,
              xlab = "", ylab = "", zlab = "",
              main = "Normal ellipsoid space")

## A cylindrical space with a decreasing variance per axis
scatterplot3d(space.maker(2500, 3, c(rnorm, rnorm, runif),
              scree = c(0.7, 0.2, 0.1)), pch = 20,
              xlab = "", ylab = "", zlab = "",
              main = "Normal cylindrical space")

## Resetting the graphic parameters
par(op)
```

>>>>

Once a matrix is generated, it is possible to apply inapplicable characters to it for increasing realism!
Inapplicable characters are commonly designated as `NA` or simply `-`.
They differ from missing characters `?` in their nature by being inapplicable rather than unknown.
For example, considering a binary character defined as "colour of the tail" with the following states "blue" and "red"; on a taxa with no tail, the character should be coded as inapplicable ("`-`") since the state of the character "colour of tail" is *known*: it's neither "blue" or "red", it's just not there!
It contrasts with coding it as missing ("`?`" - also called as ambiguous) where the state is *unknown*, for example, the taxon of interest is a fossil where the tail has no colour preserved or is not present at all due to bad conservation!

@@

## The first few lines of the resulting matrix
my_matrix_NA[1:10, 90:100]
```
<!--TG: Add inapplicable paper here -->

@@

There are many parameters that can create a "realistic" matrix (i.e. not too different from the input tree with a consistency and retention index close to what is seen in the literature) but because of the randomness of the matrix generation not all parameters combination end up creating "good" matrices.
The following parameters however, seem to generate fairly "realist" matrices with a starting coalescent tree, equal rates model with 0.85 binary characters and 0.15 three state characters, a gamma distribution with a shape parameter ($\alpha$) of 5 and no scaling ($\beta$ = 1) with a rate of 100.

@@

<!-- ## Simulating continuous morphological data -->
<!-- TG: Add tutorial for evolving "morpho" shapes -->

## Simulating multidimensional spaces
Another way to simulate data is to directly simulate an ordinated space with the `space.maker` function.
This function allows users to simulate multidimensional spaces with a certain number of properties.
For example, it is possible to design a multidimensional space with a specific distribution on each axis, a correlation between the axes and a specific cumulative variance per axis.
This can be useful for creating ordinated spaces for null hypothesis, for example if you're using the function `null.test` [@diaz2016global].

This function takes as arguments the number of elements (data points - `elements` argument) and dimensions (`dimensions` argument) to create the space and the distribution functions to be used for each axis.
The distributions are passed through the `distribution` argument as... modular functions!
You can either pass a single distribution function for all the axes (for example `distribution = runif` for all the axis being uniform) or a specific distribution function for each specific axis (for example `distribution = c(runif, rnorm, rgamma))` for the first axis being uniform, the second normal and the third gamma).
You can of course use your very own functions or use the ones implemented in `dispRity` for more complex ones (see below).
Specific optional arguments for each of these distributions can be passed as a list via the `arguments` argument.

Furthermore, it is possible to add a correlation matrix to add a correlation between the axis via the `cor.matrix` argument or even a vector of proportion of variance to be bear by each axis via the `scree` argument to simulate realistic ordinated spaces.

Here is a simple two dimensional example:

```{r, fig.width=8, fig.height=8, eval=TRUE}
## Graphical options
op <- par(bty = "n")

## A square space
square_space <- space.maker(100, 2, runif)

## The resulting 2D matrix
head(square_space)

## Visualising the space
plot(square_space, pch = 20, xlab = "", ylab = "", main = "Uniform 2D space")
```

Of course, more complex spaces can be created by changing the distributions, their arguments or adding a correlation matrix or a cumulative variance vector:

```{r}
## A plane space: uniform with one dimensions equal to 0
plane_space <- space.maker(2500, 3, c(runif, runif, runif),
                           arguments = list(list(min = 0, max = 0), NULL, NULL))

## Correlation matrix for a 3D space
(cor_matrix <- matrix(cbind(1, 0.8, 0.2, 0.8, 1, 0.7, 0.2, 0.7, 1), nrow = 3))

## An ellipsoid space (normal space with correlation)
ellipse_space <- space.maker(2500, 3, rnorm, cor.matrix = cor_matrix)

## A cylindrical space with decreasing axes variance
cylindrical_space <- space.maker(2500, 3, c(rnorm, rnorm, runif),
                                 scree = c(0.7, 0.2, 0.1))
```

### Personalised dimensions distributions

Following the modular architecture of the package, it is of course possible to pass home made distribution functions to the `distribution` argument.
For example, the `random.circle` function is a personalised one implemented in `dispRity`.
This function allows to create circles based on basic trigonometry allowing to axis to covary to produce circle coordinates.
By default, this function generates two sets of coordinates with a `distribution` argument and a minimum and maximum boundary (`inner` and `outer` respectively) to create nice sharp edges to the circle.
The maximum boundary is equivalent to the radius of the circle (it removes coordinates beyond the circle radius) and the minimum is equivalent to the radius of a smaller circle with no data (it removes coordinates below this inner circle radius).

```{r, fig.width=8, fig.height=8, eval=TRUE}
## Graphical options
op <- par(bty = "n")

## Generating coordinates for a normal circle with a upper boundary of 1
circle <- random.circle(1000, rnorm, inner = 0, outer = 1)

## Plotting the circle
plot(circle, xlab = "x", ylab = "y", main = "A normal circle")

## Creating doughnut space (a spherical space with a hole)
doughnut_space <- space.maker(5000, 3, c(rnorm, random.circle),
     arguments = list(list(mean = 0), list(runif, inner = 0.5, outer = 1)))
```

### Visualising the space

I suggest using the excellent `scatterplot3d` package to play around and visualise the simulated spaces:

@@

## The plane space
scatterplot3d(plane_space, pch = 20, xlab = "", ylab = "", zlab = "",
              xlim = c(-0.5, 0.5), main = "Plane space")

## The ellipsoid space
scatterplot3d(ellipse_space, pch = 20, xlab = "", ylab = "", zlab = "",
              main = "Normal ellipsoid space")

## A cylindrical space with a decreasing variance per axis
scatterplot3d(cylindrical_space, pch = 20, xlab = "", ylab = "", zlab = "",
              main = "Normal cylindrical space")
## Axes have different orders of magnitude

## Plotting the doughnut space
scatterplot3d(doughnut_space[,c(2,1,3)], pch = 20, xlab = "", ylab = "",
              zlab = "", main = "Doughnut space")
par(op)
```

### Generating realistic spaces

It is possible to generate "realistic" spaces by simply extracting the parameters of an existing space and scaling it up to the simulated space.
For example, we can extract the parameters of the `BeckLee_mat50` ordinated space and simulate a similar space.

```{r, fig.width=8, fig.height=8, eval=TRUE}
## Loading the data
data(BeckLee_mat50)

## Number of dimensions
obs_dim <- ncol(BeckLee_mat50)

## Observed correlation between the dimensions
obs_correlations <- cor(BeckLee_mat50)

## Observed mean and standard deviation per axis
obs_mu_sd_axis <- mapply(function(x,y) list("mean" = x, "sd" = y),
                         as.list(apply(BeckLee_mat50, 2, mean)),
                         as.list(apply(BeckLee_mat50, 2, sd)), SIMPLIFY = FALSE)

## Observed overall mean and standard deviation
obs_mu_sd_glob <- list("mean" = mean(BeckLee_mat50), "sd" = sd(BeckLee_mat50))

## Scaled observed variance per axis (scree plot)
obs_scree <- variances(BeckLee_mat50)/sum(variances(BeckLee_mat50))

## Generating our simulated space
simulated_space <- space.maker(1000, dimensions = obs_dimensions, 
                               distribution = obs_distributions,
                               arguments = obs_arguments,
                               cor.matrix = obs_correlations)

## Visualising the fit of our data in the space (in the two first dimensions)
plot(simulated_space[,1:2], xlab = "PC1", ylab = "PC2")
points(BeckLee_mat50[,1:2], col = "red", pch = 20)
legend("topleft", legend = c("observed", "simulated"),
        pch = c(20,21), col = c("red", "black"))
```

It is now possible to simulate a space using these observed arguments to test several hypothesis:

 * Is the space uniform or normal?
 * If the space is normal, is the mean and variance global or specific for each axis?

```{r, fig.width=8, fig.height=8, eval=TRUE}
## Measuring disparity as the sum of variance
observed_disp <- dispRity(BeckLee_mat50, metric = c(median, centroids))

## Is the space uniform?
test_unif <- null.test(observed_disp, null.distrib = runif)

## Is the space normal with a mean of 0 and a sd of 1?
test_norm1 <- null.test(observed_disp, null.distrib = rnorm)

## Is the space normal with the observed mean and sd and cumulative variance
test_norm2 <- null.test(observed_disp, null.distrib = rep(list(rnorm), obs_dim),
                        null.args = rep(list(obs_mu_sd_glob), obs_dim),
                        null.scree = obs_scree)

## Is the space multiple normal with multiple means and sds and a correlation?
test_norm3 <- null.test(observed_disp, null.distrib = rep(list(rnorm), obs_dim),
                        null.args = obs_mu_sd_axis, null.cor = obs_correlations)

## Graphical options
op <- par(mfrow = (c(2, 2)), bty = "n")
## Plotting the results
plot(test_unif, main = "Uniform (0,1)")
plot(test_norm1, main = "Normal (0,1)")
plot(test_norm2, main = paste0("Normal (", round(obs_mu_sd_glob[[1]], digit = 3),
                              ",", round(obs_mu_sd_glob[[2]], digit = 3), ")"))
plot(test_norm4, main = "Normal (variable + correlation)")

```

If we measure disparity as the median distance from the morphospace centroid, we can explain the distribution of the data as normal with the variable observed mean and standard deviation and with a correlation between the dimensions.
